Importance of background gene list for enrichment analysis

### 2023-02-23 bg2.Rmd
fix minor bugs

For this guide I will be using data from a previous analysis, which is
deposited at NCBI GEO and SRA under accession numbers: GSE109140 and SRP128998.

The data was presented in a [paper by Felisbino et al](https://pubmed.ncbi.nlm.nih.gov/33495488/)

@@

library("getDEE2")
library("DESeq2")
library("kableExtra")
library("vioplot")
library("clusterProfiler")
library("eulerr")

@@

Using RNAseq data from [DEE2](https://pubmed.ncbi.nlm.nih.gov/30942868/),
which can be accessed from dee2.io

@@

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP128998")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)

# vpa subset no filtering
ss1 <- ss[grep("low",ss$Experiment_title),]
ss1$vpa <- grepl("VPA",ss1$Experiment_title)

ss1 %>% 
  kbl(caption="sample sheet for VPA treatment in hepatocytes") %>%
  kable_paper("hover", full_width = F)

mx1 <- mx[,which(colnames(mx) %in% ss1$SRR_accession)] 
dim(mx1)

@@

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx1)),horiz=TRUE,las=1,main="number of reads per sample in SRP128998")

@@

# look at the distribution of counts
l1 <- lapply(1:ncol(mx1),function(i) log10( mx1[,i] +1 ))
names(l1) <- colnames(mx1)
l1 <- rev(l1)

hist(l1[[6]],main="Distribution of read counts in SRR6467485",
  xlab="log10(readcounts+1)",breaks=30)

@@

# DESeq2 without any fitering
dds <- DESeqDataSetFromMatrix(countData = mx1 , colData = ss1, design = ~ vpa )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),mx1)
dge1<-as.data.frame(zz[order(zz$pvalue),])

head(dge1,10) %>% 
  kbl(caption="Top DE genes for VPA treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

sig1 <- subset(dge1,padj<=0.05)
message("number of significant genes")
nrow(sig1)

plot(rowMeans( log10(dge1[,7:12]+1) ),dge1$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  xlab="log10(mean read count)",ylab="log2(fold change)")

points(rowMeans( log10(sig1[,7:12]+1) ),sig1$log2FoldChange,
  cex=0.6,pch=19,col="red")

@@

dge1up <- rownames(subset(dge1,padj<=0.05 & log2FoldChange>0))
dge1up <- unique(sapply(strsplit(dge1up," "),"[[",2))

dge1dn <- rownames(subset(dge1,padj<=0.05 & log2FoldChange<0))
dge1dn <- unique(sapply(strsplit(dge1dn," "),"[[",2))

dge1bg <- rownames(dge1)
dge1bg <- unique(sapply(strsplit(dge1bg," "),"[[",2))

@@

sig1 <- sig1[order(sig1$baseMean),]
head(sig1,10) %>%
  kbl(caption="lowest basemean genes") %>%
  kable_paper("hover", full_width = F)

@@

mx2 <- mx1[which(rowMeans(mx1)>=10),]
dim(mx2)
dds <- DESeqDataSetFromMatrix(countData = mx2 , colData = ss1, design = ~ vpa )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),mx2)
dge2<-as.data.frame(zz[order(zz$pvalue),])

head(dge2,10) %>%
  kbl(caption="Top DE genes for VPA treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

sig2 <- subset(dge2,padj<=0.05)
nrow(sig2)

plot(log10(dge2$baseMean),dge2$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  xlab="log10(base mean)",ylab="log2 fold change",
  main="strict filter")

points(log10(sig2$baseMean),sig2$log2FoldChange,
  cex=0.6,pch=19,col="red")

dge2up <- rownames(subset(dge2,padj<=0.05 & log2FoldChange>0))
dge2up <- unique(sapply(strsplit(dge2up," "),"[[",2))

dge2dn <- rownames(subset(dge2,padj<=0.05 & log2FoldChange<0))
dge2dn <- unique(sapply(strsplit(dge2dn," "),"[[",2))

dge2bg <- rownames(dge2)
dge2bg <- unique(sapply(strsplit(dge2bg," "),"[[",2))

@@

ora2_up <- as.data.frame(enricher(gene = dge2up , 
  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_up$geneID <- NULL
ora2_ups <- rownames(subset(ora2_up,p.adjust<0.05))

@@

ora2_up$es <- gr/br
ora2_up <- ora2_up[order(-ora2_up$es),]
head(ora2_up)

@@

ora2_dn <- as.data.frame(enricher(gene = dge2dn ,
  universe = dge2bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_dn$geneID <- NULL
ora2_dns <- rownames(subset(ora2_dn,p.adjust<0.05))

@@

ora2_dn$es <- gr/br
ora2_dn <- ora2_dn[order(-ora2_dn$es),]
head(ora2_dn)

@@

ora1_up <- as.data.frame(enricher(gene = dge1up ,
  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_up$geneID <- NULL
ora1_ups <- rownames(subset(ora1_up,p.adjust<0.05))

@@

ora1_up$es <- gr/br
ora1_up <- ora1_up[order(-ora1_up$es),]
head(ora1_up)

@@

ora1_dn <- as.data.frame(enricher(gene = dge1dn ,
  universe = dge1bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_dn$geneID <- NULL
ora1_dns <- rownames(subset(ora1_dn,p.adjust<0.05))

@@

ora1_dn$es <- gr/br
ora1_dn <- ora1_dn[order(-ora1_dn$es),]
head(ora1_dn)

@@

```{r,venn1}

@@

par(mfrow=c(2,1))
plot(euler(v1),quantities = TRUE)
plot(euler(v2),quantities = TRUE)
dev.off()

@@

@@

>>>>

For this guide I will be using data from a previous analysis, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101.

The data was presented in a [paper by Lund et al, 2014](https://pubmed.ncbi.nlm.nih.gov/25315154/)

The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

@@

suppressPackageStartupMessages({
  library("getDEE2")
  library("DESeq2")
  library("kableExtra")
  library("vioplot")
  library("clusterProfiler")
  library("eulerr")
})

@@

Using RNAseq data from [DEE2](https://pubmed.ncbi.nlm.nih.gov/25315154/),
which can be accessed from dee2.io

@@

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

@@

par(mar=c(5,7,5,1))
barplot(rev(colSums(mx)),horiz=TRUE,las=1,main="number of reads per sample in SRP038101")

@@

# look at the distribution of counts
l <- lapply(1:ncol(mx),function(i) log10( mx[,i] +1 ))
names(l) <- colnames(mx)
l <- rev(l)

blah()
hist(l[[6]],main="Distribution of read counts in SRR1171523",
  xlab="log10(readcounts+1)",breaks=30)

@@

# DESeq2 without any fitering
dds <- DESeqDataSetFromMatrix(countData = mx , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),mx)
de <- as.data.frame(zz[order(zz$pvalue),])

head(de,10) %>%
  kbl(caption="Top DE genes for Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

DET=nrow(mx)
NSIG=nrow(sig)
NUP=nrow(subset(sig,log2FoldChange>0))
NDN=nrow(subset(sig,log2FoldChange<0))

HEADER=paste(DET,"genes included;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(de$baseMean) ,de$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="no detection filter",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sig$baseMean) ,sig$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

@@

deup <- rownames(subset(de,padj<=0.05 & log2FoldChange>0))
deup <- unique(sapply(strsplit(deup," "),"[[",2))

dedn <- rownames(subset(de,padj<=0.05 & log2FoldChange<0))
dedn <- unique(sapply(strsplit(dedn," "),"[[",2))

all <- rownames(de)
all <- unique(sapply(strsplit(all," "),"[[",2))

@@

sig <- sig[order(sig$baseMean),]

head(sig,10) %>%
  kbl(caption="lowest basemean genes") %>%
  kable_paper("hover", full_width = F)

@@

mxf <- mx[which(rowMeans(mx)>=10),]
dim(mxf)
dds <- DESeqDataSetFromMatrix(countData = mxf , colData = ss, design = ~ trt )
res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <-cbind(as.data.frame(z),mxf)
def <-as.data.frame(zz[order(zz$pvalue),])

head(def,10) %>%
  kbl(caption="Top DE genes for Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

sigf <- subset(def,padj<=0.05)

DET=nrow(mxf)
NSIG=nrow(sigf)
NUP=nrow(subset(sigf,log2FoldChange>0))
NDN=nrow(subset(sigf,log2FoldChange<0))

HEADER=paste(DET,"detected genes;",NSIG,"w/FDR<0.05;",NUP,"up;",NDN,"down")

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="strict filter",
  xlab="log10(basemean)",ylab="log2(fold change)")

points(log10(sigf$baseMean) ,sigf$log2FoldChange,
  cex=0.6,pch=19,col="red")

mtext(HEADER)

defup <- rownames(subset(def,padj<=0.05 & log2FoldChange>0))
defup <- unique(sapply(strsplit(defup," "),"[[",2))

defdn <- rownames(subset(def,padj<=0.05 & log2FoldChange<0))
defdn <- unique(sapply(strsplit(defdn," "),"[[",2))

bg <- rownames(def)
bg <- unique(sapply(strsplit(bg," "),"[[",2))

@@

ora2_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_up$geneID <- NULL
ora2_up <- subset(ora2_up,p.adjust<0.05 & Count >=10)
ora2_ups <- rownames(ora2_up)

@@

ora2_up$es <- gr/br
ora2_up <- ora2_up[order(-ora2_up$es),]
ora2_up$Description=NULL

head(ora2_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

@@

ora2_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora2_dn$geneID <- NULL
ora2_dn <- subset(ora2_dn,p.adjust<0.05 & Count >=10)
ora2_dns <- rownames(ora2_dn)

@@

ora2_dn$es <- gr/br
ora2_dn <- ora2_dn[order(-ora2_dn$es),]
ora2_dn$Description=NULL

head(ora2_dn) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (filtered)") %>%
  kable_paper("hover", full_width = F)

@@

ora1_up <- as.data.frame(enricher(gene = defup ,
  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_up$geneID <- NULL
ora1_up <- subset(ora1_up,p.adjust<0.05 & Count >=10)
ora1_ups <- rownames(ora1_up)

@@

ora1_up$es <- gr/br
ora1_up <- ora1_up[order(-ora1_up$es),]
ora1_up$Description=NULL

head(ora1_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

@@

ora1_dn <- as.data.frame(enricher(gene = defdn ,
  universe = all,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

ora1_dn$geneID <- NULL
ora1_dn <- subset(ora1_dn,p.adjust<0.05 & Count >=10)
ora1_dns <- rownames(ora1_dn)

@@

ora1_dn$es <- gr/br
ora1_dn <- ora1_dn[order(-ora1_dn$es),]
ora1_dn$Description=NULL

head(ora1_dn) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in Aza treatment (unfiltered)") %>%
  kable_paper("hover", full_width = F)

@@

```{r,venn1,fig.width=6,fig.height=6}

@@

par(mfrow=c(2,1))
plot(euler(v1),quantities = TRUE)
plot(euler(v2),quantities = TRUE)

Jaccard index

```{r,jaccard}

jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}

correct <- c(ora2_ups,ora2_dns)
incorrect <- c(ora1_ups,ora1_dns)

jaccard(correct,incorrect)

@@

For reproducibility

<br><details><summary><b>Click HERE to show session info</b></summary><br><p>

@@

</details>
